---
title: "Suicide Rates Project"
author: "Nick Wawee"
date: "10/16/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

## Libraries 

```{r setup, message = F, warning = F}
rm(list=ls())
library(car)
library(ggplot2)
library(ggrepel)
library(reshape2)
library(ggpubr)
library(dplyr)
library(limma)
library(MASS)
knitr::opts_chunk$set(echo = TRUE)
options(width = 60)
plot_opts =   theme_bw()+
  theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black") , axis.title.x = element_text(face='bold', size = 14), axis.title.y = element_text(face='bold', size = 14), plot.title = element_text(face='bold', size = 18, hjust=0.5))
```


## Functions

```{r functions}
plotdists = function(df, path){
  plotlist = list()
  for (col in colnames(df)){
    x = df[,col]
    if (class(x)=='factor'){
      dfnew = data.frame(col = factor(), count = integer())
      for (level in levels(x)){
        count = length(which(x==level))
        dfnew = rbind(data.frame(col= level, count = count), dfnew)
      }
      dfnew$col <- factor(dfnew$col, levels = dfnew$col[order(dfnew$count)])
      p = ggplot(dfnew, aes(x=col, y=count))+
        geom_bar(stat= 'identity')+
        plot_opts+
        labs(x = col, y = 'Count', title = paste(col, "Distribution"))+
        geom_text(aes(label = count), vjust = -0.3) 
      plotlist[[col]] = p
    }else if (class(x) == 'numeric' | class(x) == 'integer'){
      dfnew =data.frame(col = class(x))
      histinfo = hist(x = x , breaks='Scott', plot = F)
      p = ggplot(as.data.frame(x), aes(x=x))+
        geom_histogram(bins = length(histinfo$breaks))+plot_opts+
        #geom_density(aes(y=..count..), size = 2)+plot_opts+
        geom_vline(aes(xintercept = median(x)), 
             linetype = "dashed", size = 0.6)+
        labs(x = col, y = 'Count', title = paste(col, 'Distribution'))
        
      plotlist[[col]] = p 
    }
  }
  pfinal =ggarrange(plotlist = plotlist)
  ggsave(path, pfinal, height=2.5*length(plotlist) , width=2.5*length(plotlist), units="in", limitsize = FALSE, dpi=300)
  return(pfinal)
}

Assumption_Check = function(l_m, outp){
  df = l_m[['model']]
  df$residuals = l_m[["residuals"]]
  df = df[,-which(colnames(df)==l_m[["call"]][["formula"]][[2]])]
  #Assumption 1 - linearity check
  a1pls = list()#assumption 1 plotlist
  for (col in colnames(df)[-which(colnames(df)=='residuals')]){
    dfnew = data.frame(x = df[,col], y = df$residuals)
    a1pls[[col]] = ggplot(data = dfnew, aes(x =x, y=y))+
        geom_jitter()+plot_opts+labs(x = col, y = 'Residuals')+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))+
      geom_hline(yintercept=0,linetype="dashed",color = "red", size=1)
  }
  p1 = ggarrange(plotlist= a1pls)
  p1 = annotate_figure(p1,text_grob("Linearity Checks", color = "red", face = "bold", size = 20))
  ggsave(filename = paste(outp,'linearity.png',sep=""), plot = p1, dpi = 600)
  
  #Assumption 2 - Bias and Scedasticity 
  df2 = data.frame(Fit = l_m$fitted.values, Residuals = l_m$residuals)
  p2 = ggplot(data = df2, aes(x = Fit, y = Residuals))+
    geom_point()+plot_opts+geom_hline(yintercept=0,linetype="dashed",color = "red", size=1)+labs(title = 'Bias and Scedasticity Check')+theme(plot.title = element_text(color = 'red'))
  ggsave(filename = paste(outp,'bias_sced.png',sep=""), plot = p2, dpi = 600)
  
  #Assumption 3 - Correlation in Errors
  a3pls = list()
  for (n in colnames(df[,-which(colnames(df)=='residuals')])){
    dfnew = data.frame(y = df$residuals[order(df[,n])], x = 1:nrow(df))
    a3pls[[n]] = ggplot(data = dfnew, aes(x = x, y = y))+geom_jitter()+
      plot_opts+labs(title = paste('Sorted By:', n), x = 'Index', y='Residuals')
  }
  p3 = ggarrange(plotlist = a3pls)
  p3 = annotate_figure(p3,text_grob("Error Independence Check", color = "red", face = "bold", size = 20))
  ggsave(filename = paste(outp,'inderror.png',sep=""), plot = p3, dpi = 600)

  #Assumption 4 - Normality of Residuals
  shapres = shapiro.test(l_m$residuals)
  p4 = ggplot(df)+
    geom_qq(aes(sample = residuals))+geom_qq_line(aes(sample= residuals))+
    plot_opts+ labs(title = paste('Normality of Residuals\n', 'Shaprio Wilks Results: W = ', as.character(round(shapres$statistic,3)), ', p = ', as.character(round(shapres$p.value,5))), x = 'Theoretical Values', y = 'Sample Values')+ theme(plot.title = element_text(color = 'red'))
  ggsave(filename = paste(outp,'normres.png',sep=""), plot = p4, dpi = 600)
  
  pfinal = ggarrange(plotlist = list(p1, p2, p3, p4))
  pfinal= annotate_figure(pfinal,text_grob("Model Assumption Check", face = "bold", size = 26))
  ggsave(filename = paste(outp,'all_assum.png',sep=""), plot = pfinal, dpi = 600, width = 8, height = 10, units = 'in')
  return(pfinal)
}

infl_analysis = function(l_m, df){
  k = length(l_m$coefficients)-1
  n = nrow(df)
  row_num = 1:n
  #response_v = df[colnames(df)==l_m$terms[[2]]] #use if you would like to change the labels of the points to the response variable rather than observation number
  #Leverage points
  hatdf = data.frame(Values = hatvalues(l_m), Row_Num = row_num, Type = rep('Hat Values', length(row_num)), Point_Type = rep('Leverage', length(row_num)), Bound1 = 2*(k+1)/n, Bound2 = 2*(k+1)/n)
  hatdf$Label = NA
  inds = which(hatvalues(l_m)>2*(k+1)/n)
  if(length(inds)!= 0){hatdf$Label[inds] = row_num[inds]}
  #Outliers
  instdf = data.frame(Values = rstandard(l_m), Row_Num = row_num, Type = rep('Internally Standardized Residuals', length(row_num)), Point_Type = rep('Outlier', length(row_num)), Bound1 = 3, Bound2 = -3)
  instdf$Label = NA
  inds = which(rstandard(l_m) > 3 | rstandard(l_m) < -3)
  if(length(inds)!=0){instdf$Label[inds] = row_num[inds]}
  
  extdf = data.frame(Values = rstudent(l_m), Row_Num = row_num, Type = rep('Externally Standardized Residuals', length(row_num)), Point_Type = rep('Outlier', length(row_num)), Bound1 = 3, Bound2 = -3)
  extdf$Label = NA
  inds = which(rstudent(l_m) > 3 | rstudent(l_m) < -3)
  if(length(inds)!=0){extdf$Label[inds] = row_num[inds]}
  
  #Influential
  dfitsdf = data.frame(Values = dffits(l_m), Row_Num = row_num, Type = rep('DEFFITS', length(row_num)),Point_Type = rep('Influential', length(row_num)), Bound1 = 2*sqrt((k+2)/(n-k-2)), Bound2 = -2*sqrt((k+2)/(n-k-2)))
  dfitsdf$Label = NA
  inds = which(dffits(l_m) > 2*sqrt((k+2)/(n-k-2)) | dffits(l_m) < -2*sqrt((k+2)/(n-k-2)))
  if(length(inds)!=0){dfitsdf$Label[inds] = row_num[inds]}
  
  cddf = data.frame(Values = cooks.distance(l_m), Row_Num = row_num, Type = rep("Cook's Distance", length(row_num)),Point_Type = rep('Influential', length(row_num)), Bound1 = 1, Bound2 = 1)
  cddf$Label = NA
  inds = cooks.distance(l_m) > 1
  if(length(inds)!=0){cddf$Label[inds] = row_num[inds]}
  
  cvdf = data.frame(Values = covratio(l_m), Row_Num = row_num, Type = rep("Covariance Ratio", length(row_num)),Point_Type = rep('Influential', length(row_num)), Bound1 = 1 + 3*(k+1)/n, Bound2 = 1 - 3*(k+1)/n)
  cvdf$Label = NA
  inds = covratio(l_m) > 1 + 3*(k+1)/n | covratio(l_m) < 1 - 3*(k+1)/n
  if(length(inds)!=0){cvdf$Label[inds] = row_num[inds]}
  
  ret_df = rbind(hatdf, instdf, extdf, dfitsdf, cddf, cvdf)
  return(ret_df)
}
```

## Loading

```{r loading}
df = read.csv("../Data/master.csv", stringsAsFactors = T)
str(df)
```

## Cleaning 

Below will examine the na values in each column:

```{r description}
for (col in colnames(df)){
  print(length(which(is.na(df[,col]))))
}
```

It appears that the HDI for year column has the majority of NAs and will be discluded from the remaining of the analysis. The data will be filtered for the United States only.

```{r removeHDI, echo = F}
df = df[,-9]
df = df[df$country == 'United States',]
df = df[, c(-1,-5,-6,-8, -11)]
df$gdp_for_year.... = as.numeric(gsub(",", "", df$gdp_for_year....))
str(df)
df = df[,c(-1,-5)]#for final mlr model
```

## Distributions 

```{r dists, fig.height = 6, fig.width = 6}
colnames(df) = c('Sex', 'Age', 'Suicide_Rate', 'GDP_per_Capita')
df$Age <- relevel(df$Age, ref = '5-14 years')
df$Sex = factor(paste(toupper(strsplit2(df$Sex, split ="")[,1])))
path = "../Plots/distplots.png"
plotdists(df, path)
```

## Model Fitting

```{r mlr2}
mlr2 = lm(Suicide_Rate ~ Sex*Age + GDP_per_Capita, data = df)
summary(mlr2)
```

## Scatter Plot

```{r scatter plot}
outp = "../Plots/"
p = ggplot(data = df, aes(x = GDP_per_Capita, y = Suicide_Rate))+
  geom_point(aes(color = Age), alpha = 0.5)+plot_opts+facet_wrap(~Sex)+
  labs(x = 'GDP per Capita', y = 'Suicide Number / 100k Population', title = 'Suicide Rate in US from 1985 - 2016')
p
ggsave(filename = paste(outp,'scatter.png',sep=""), plot = p, dpi = 600, width = 8, height = 4, units = 'in')
```

## Assumption Check 1

```{r assumption check usage, fig.width=12, fig.height=12, units = 'in'}
pf = Assumption_Check(mlr2, outp)
pf
```

## Influential Point Analysis

```{r influential analysis, warning = F, message = F, fig.width=12, fig.height=6, units = 'in'}
ret_df = infl_analysis(mlr2, df =df)
ret_df = cbind(ret_df, df)
p = ggplot(data= ret_df, aes(x= Row_Num, y = Values))+
  geom_point(aes(color = Age, shape = Sex))+
  facet_wrap(~Type, scales = "free_y")+plot_opts+geom_line(aes(y=Bound1))+geom_line(aes(y=Bound2))+
  geom_label_repel(aes(label=Label))+
  labs(title = 'Influential Point Analysis', x = 'Observation Number')
p
ggsave(filename = paste(outp,'influential.png',sep=""), plot = p, dpi = 600, width = 12, height = 6, units = 'in')
```

## BoxCox Transformation

```{r boxcox}
bc = boxcox(mlr2, data = df)
p = bc$x[which.max(bc$y)]
df$Suicide_Rate = df$Suicide_Rate**p
```

## Transformed Response Variable

```{r mlr3, echo =F}
mlr3 = lm(Suicide_Rate ~ Sex*Age + GDP_per_Capita, data = df)
summary(mlr3)
```

## Assumption Check 2

```{r assumption check usage2, fig.width=12, fig.height=12, units = 'in'}
outp = "../Plots/trans_"
pf = Assumption_Check(mlr3, outp)
pf
```

## Influential Point Analysis 2

```{r influential analysis2, warning = F, message = F, fig.width=12, fig.height=6, units = 'in'}
ret_df = infl_analysis(mlr3, df =df)
ret_df = cbind(ret_df, df)
p = ggplot(data= ret_df, aes(x= Row_Num, y = Values))+
  geom_point(aes(color = Age, shape = Sex))+
  facet_wrap(~Type, scales = "free_y")+plot_opts+geom_line(aes(y=Bound1))+geom_line(aes(y=Bound2))+
  geom_label_repel(aes(label=Label))+
  labs(title = 'Influential Point Analysis', x = 'Observation Number')
p
ggsave(filename = paste(outp,'influential.png',sep=""), plot = p, dpi = 600, width = 12, height = 6, units = 'in')
```

## Transformed Response Plot

```{r final plot}
p = ggplot(data = df, aes(x = GDP_per_Capita, y = Suicide_Rate))+
  geom_point(aes(color = Age), alpha = 0.5)+plot_opts+facet_wrap(~Sex)+
  labs(x = 'GDP per Capita', y = '[Suicide Number / 100k Population]^0.141', title = 'Suicide Rate in US from 1985 - 2016')
p
ggsave(filename = paste(outp,'scatter.png',sep=""), plot = p, dpi = 600, width = 10, height = 5, units = 'in')
```